---
title: "Homeless Populations in Toronto"
author: "Jack McKay"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2
toc: FALSE
abstract: "Shelter System Flow data was collected by the City of Toronto Open Portal, in order to record what populations are most affected by homelessness in Toronto as well as monitor how the City’s Shelter System is functioning. Shelter usage is an important factor in determining the livability and accessibility of a city, and monitoring trends and fluctuations can allow us to better cater the services provided to the populations of people that need it the most."
thanks: 'Code and data are available at [github.com/jmacattack27/ModellingHomelessPopulations](https://github.com/jmacattack27/ModellingHomelessPopulations).'
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
#install.packages("opendatatoronto")
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(knitr)
library(janitor)
library(lubridate)
library(tibble)
library(dplyr)
library(scales)

```

# Introduction

During the 2021 elections, there was one issue that every major party unanimously agreed upon; Canada is in the midst of a severe housing crisis, and the cost of home ownership must be lowered. Since the early 2000s, Canada experienced an intense surge in housing prices. Known as the Canadian Housing Crisis, the City of Toronto has been one of the driving forces in this surge. In the last two decades, home prices in Canada have increased by 375%, while prices in Toronto have soared over 450% (@clarke). One of the driving factor of this is the increasing commodification of houses, wherein foreign investors purchase available real estate and hold it as an investment. This greatly restricts the supply of available housing in Toronto, and has a myriad of negative consequences for citizens. One such consequence is the rapidly growing homeless population in Toronto. As is commonplace with socioeconomic issues, the ones most affected by this crisis are those who are most the most economically vulnerable. Increasingly, the trend of housing prices grows more and more detached from the trends of income prices, leading housing to become less affordable to a greater number of people as this crisis progresses. The problem of Toronto's growing homeless population has reared it's ugly head at the City over the last 2 years, as Covid-19 fears and lack of adequate social distancing in shelters caused a huge outflow of people onto the streets. This made visible to all the increasing urban destitution that has been largely driven by the housing crisis. Indeed, as this problem becomes unavoidable to even the most neglecting lawmakers, the City has responded with increasingly aggressive and vacuous measures; in the summer of 2021, the city spent $2 million on clearing homeless encampments (@draaisma). Rather than funding the swift and often brutal evictment of homeless people by overpaid and overequipped police officers, the City of Toronto must instead invest in its shelter system, and improve services aimed at eliminating homelessness at its roots. 


As the homeless populace of Toronto grows, so too does the importance of the Toronto Shelter System, a series of services and shelters operated and funded by the City dedicated to people experiencing homelessness. Indeed, homeless shelters play a critical role in determining the livability of a city; this is especially true with a city as large and populous as Toronto. Not only do shelters provide homeless citizens with basic necessities such as food, water and safety, but they also provide access to medical care, social workers and housing workers. This means that shelters provide both temporary refuge and necessities, as well as ways to escape the harrowing cycle of homelessness. 

This report aims to identify and highlight the areas of the Toronto Shelter System that are most in need of funding, through analyzing trends in the homeless population as a whole, as well as the sub-populations of people that make up the aggregate.  


# Data


## Data Source ##

The data used in this report was collected through the Shelter Management Information System (SMIS), an information system used by the City of Toronto to monitor and manage shelters, warming centres, and other allied services that are funded by the City. This data was then collated and stored on the City of Toronto Open Portal, from where it was pulled for this report. 


## Data Collection ##

This dataset contains information on the daily occupancy and capacity of Toronto’s shelter system since January 2020. SMIS records the number of unique people who have used the shelter system at least one time per month for the last 3 months, who have not been discharged to permanent housing. Data is recorded daily at 4 a.m., and includes occupancy, vacancy and capacity data for each service. This data is recorded based on the capacity type for each program, either bed based capacity (where occupancy is measured based at the bed level) or room based capacity (where occupancy is measured at the room level). Note that in room based capacity, entire households share a single room, so the total number of individuals served is often greater than both capacity and occupancy.  

SMIS records the age and gender of individuals, as well as if they: have used a shelter service in the last 3 months (Actively Homeless), have used the shelter system previously but not in the last 3 months and have now returned (Returned to Shelter), have previously been discharged to permanent housing and have used the shelter system in the recording month (Returned from Housing), are using a shelter for the first time (Newly Identified), or if they have previously used a shelter service, but not in the last 3 months (No Recent Shelter Use). Additionally, SMIS records data on various sub-populations, including: Chronically Homeless, Families, Single Adult, Unaccompanied Youth, Refugees and Indigenous. Someone who is Chronically Homeless is defined as having recorded a minimum of 180 overnight stays in the past year (365 days); or having recurrent overnight stays over the past three years with a cumulative duration of at least 546 nights. 


## Data Analysis ##

The first aspect of the data I investigated relates to the attrition rate of housing, ie; the likelihood that a homeless person moved to permanent housing does not return to the shelter. Figure \@ref(fig:housingattrition) graphs the number of people moved into permanent housing against the number of people who have returned to a shelter from after being moved to permanent housing. There has been a sharp decrease in the number of people being moved to housing since the beginning of 2020, while the number of people returning to shelters after being placed in permanent housing has remained relatively stable. This indicates a sharp decrease in the attrition rates of housing placements, which is one of the primary gauges of how successful Toronto's homeless services are at eliminating homelessness. 


```{r, include=FALSE}
populations <- read.csv(here::here("inputs/data/raw_data.csv"))

populations$date.mmm.yy. <- Map(paste, populations$date.mmm.yy., "-01", sep="")
populations$date.mmm.yy. <- strptime(populations$date.mmm.yy., "%b-%y-%d")
populations$date.mmm.yy. <- as.Date(populations$date.mmm.yy., format = "%Y-%m-%d")
names(populations)[2] <- "Date"

homeless_housing <-
  populations %>% 
  
  filter(population_group == 'All Population') %>%   # Filter data so that we only look at the full population for each month
  
  select(Date, returned_from_housing, moved_to_housing) %>% 
  
  mutate(Attrition = percent(1 - returned_from_housing / moved_to_housing)) %>% 
  
  rename("Returned from Housing" = returned_from_housing,   # Rename columns so that table reads nicer
         "Moved to Housing" = moved_to_housing)
```


```{r housingattrition, fig.cap = 'Moved to Housing vs Returned from Housing', fig.width = 8, fig.height = 2.5, echo = FALSE, fig.align = "center", fig.pos = 'h'}

graph_housing <-
  homeless_housing %>% 
  tidyr::gather("Moved to Housing vs Returned from Housing", "Population", 2:3, na.rm = FALSE, convert = FALSE) 

graph_housing %>% 
  ggplot(aes(x = Date, y = Population, color = `Moved to Housing vs Returned from Housing`, group = 2)) +
  geom_point() +
  geom_line(aes(group = `Moved to Housing vs Returned from Housing`))
```


The next aspect of the data I analyzed was the composition of the total homeless population and its fluctuations over the last two years. If applicable, the Shelter Intake Management System further classifies homeless people into the following sub-groups:

  * Chronic: Refers to chronic homelessness, which is defined as someone who has recorded a minimum of 180 overnight stays in the              past year, or has recorded a cumulative 546 overnight stays in the past 3 years.
  
  * Families: Individuals staying in a family designated overnight service
  
  * Youth: Individuals that are between 16 and 24 years of age by the end of the reporting month, and who are not members of a               family, as defined above. 
  
  * Single Adult: Individuals that are neither classified as Youth nor as part of a family, as defined above.
  
  * Refugees: Individuals that either identify as refugees when registering with a shelter system, or who register into a program               designated for refugees. When registering a family, all members of the family are registered in the same way as the               household head.  
  
  * Non-Refugees: People who are not classified as refugees.
  
  * Indigenous: A person who has self-identified as First Nations, Metis or Inuit in at least one of the intakes completed by SMIS
  
Identifying the which groups are most prevalent in shelters can allow for better allocation of funds in order to maximize the benefits to the populations most in need. Note that people only started being classified as Indigenous starting in October 2020, and it took a brief period for the data to stabilize. As a result, data on Indigenous people only began to be included on January 2021. Figure \@ref(fig:popcomposition) shows the proportion of the total homeless population accounted for by these sub-groups. Here we can see that since January 2020, the proportion of non-refugees has increased from 70% to almost 85%, while the proportion of chronically homeless people has increased by about 10%, from 35% to over 45%. Notably, the youth population has stayed quite consistent at about 10% of the total homeless population, while the proportion of single adults has increased from about 63% to over 70%. 

```{r popcomposition, fig.cap = 'Total Population Composition Since 2020', fig.width = 6, fig.height = 3, echo = FALSE, fig.align = "center", fig.pos = 'h'}
homeless_comp <-
  populations %>% 
  filter(population_group != "All Population") %>% 
  select(population_group, Date, actively_homeless, population_group_percentage)


homeless_comp %>% 
  ggplot(aes(x = Date, y = population_group_percentage, color = population_group, group = 2)) +
  geom_point() +
  geom_line(aes(group = population_group)) +
  labs(title = "Total Population Composition Over Since 2020") +
  ylab("Percentage of Total Population") +
  scale_y_continuous(breaks = pretty(homeless_comp$population_group_percentage, n = 10))
```

# Results

jabbadoo

# Discussion


bazoo 


## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


